---
title: "Dataframes - In Class lab"
author: "Dr. Shoemaker"
date: "DATA 2401"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dplyr)
library(palmerpenguins) 
data(package = 'palmerpenguins')
```

# Working with loaded in csv files


Use the `read.csv()` function to read the data from the `gates_donations.csv` file (available on Canvas, make sure you save it in the project repository folder) into a variable called `grants` 

```{r}
## if you save it in the same folder as this file this should work. 
## But you may need edit the string to reflect the actual file name: 

grants <- read_csv("gates_donations.csv")


head(grants)
```

Because you're in an RProject, you don't need to set your working directory in RStudio! The working directory is automatically set to the repository location. (You can use `getwd()` to confirm this)

- Use the glimpse function to look at the loaded data

```{r}
glimpse(grants)

```


- Create a variable `org` that contains the `organization` column of the dataset

```{r}
org <- grants$organization

```


- Confirm that the "organization" column is a vector using the `is.vector()` function. (This is a useful debugging tip if you hit errors later!)

```{r}
is.vector(org)

```


Now you can ask some interesting questions about the dataset:

- What was the mean grant value?

```{r}
mean(grants$total_amount)
```


- What was the dollar amount of the largest grant?

```{r}
max(grants$total_amount)
```


- What was the dollar amount of the smallest grant?

```{r}
min(grants$total_amount)
```


- Which organization received the largest grant?

```{r}
org[which(grants$total_amount == max(grants$total_amount))]

```


- Which organization received the smallest grant?

```{r}
org[which(grants$total_amount == min(grants$total_amount))]
```

- How many grants were awarded in 2010?

```{r}
length(grants$start_year[grants$start_year == 2010])
```

- How much was awarded in 2010? 

```{r}
sum(grants$total_amount[grants$start_year == 2010])
```


# Working with built-in data sets

Load R's "USPersonalExpenditure" dataset using the `data()` function

This will produce a data frame called `USPersonalExpenditure`
```{r}
data("USPersonalExpenditure")
```

The variable `USPersonalExpenditure` is now accessible to you. Unfortunately, it's not a data frame (it's actually a matrix, you can find that out using) 

Test this using the `is.data.frame()` function:

```{r}
is.data.frame(USPersonalExpenditure)
```

Luckily, you can pass the USPersonalExpenditure variable as an argument to the `data.frame()` function to convert it a data frame.

Do this, storing the result in a new variable

```{r}
USPersonalExpenditure <- data.frame(USPersonalExpenditure)
```


What are the current column names of your dataframe?
```{r}
colnames(USPersonalExpenditure)
```


Consider: why are they so strange? (Think about whether you could use a number like 1940 with dollar notation!)

What are the current row names of your dataframe?
```{r}
rownames(USPersonalExpenditure)
```


Add a column "Category" to your data frame that contains the rownames

```{r}
USPersonalExpenditure$Category <- rownames(USPersonalExpenditure)
USPersonalExpenditure
```

Consider how this data isn't tidy. Use the `pivot_longer()` function from last class to make the data tidy. 

```{r}
USPersonalExpenditure<- pivot_longer(USPersonalExpenditure, c("X1940", "X1945", "X1950", "X1955", "X1960"), names_to="Year", values_to="Money_Spent")
USPersonalExpenditure
```


How much money was spent on personal care in 1940?
```{r}
USPersonalExpenditure$Money_Spent[(USPersonalExpenditure$Year == "X1940") & (USPersonalExpenditure$Category == "Personal Care")]
```


How much money was spent on Food and Tobacco in 1960?
```{r}
USPersonalExpenditure$Money_Spent[(USPersonalExpenditure$Year == "X1960") & (USPersonalExpenditure$Category == "Food and Tobacco")]

```


What was the highest expenditure category in 1960?
```{r}
expenditure_sums <- list(
  Med_Health = USPersonalExpenditure$Money_Spent[(USPersonalExpenditure$Category == "Medical and Health") & (USPersonalExpenditure$Year == "X1960")],            
  Personal_care = USPersonalExpenditure$Money_Spent[(USPersonalExpenditure$Category == "Personal Care") & (USPersonalExpenditure$Year == "X1960")], 
  Food_Tobacco = USPersonalExpenditure$Money_Spent[(USPersonalExpenditure$Category == "Food and Tobacco") & (USPersonalExpenditure$Year == "X1960")], 
  Household_op = USPersonalExpenditure$Money_Spent[(USPersonalExpenditure$Category == "Household Operation") & (USPersonalExpenditure$Year == "X1960")], 
  Priv_Ed = USPersonalExpenditure$Money_Spent[(USPersonalExpenditure$Category == "Private Education") & (USPersonalExpenditure$Year == "X1960")]
)

expenditure_sums
which.max(expenditure_sums)

max_spend <- max(USPersonalExpenditure[USPersonalExpenditure$Year == 'X1960', 'Money_Spent'])
USPersonalExpenditure$Category[USPersonalExpenditure$Money_Spent == max_spend] 

USPersonalExpenditure %>% group_by(Year) %>% slice_max(Money_Spent)
```


Define a function `lowest_category` that takes in a year as a parameter, and returns the lowest spending category of that year
```{r}
lowest_category <- function(year) {
  expenditure_sums1 <- list(
  Med_Health = sum(USPersonalExpenditure$Money_Spent[(USPersonalExpenditure$Category == "Medical and Health") & (USPersonalExpenditure$Year == paste0("X", year))]),  
  Personal_care = sum(USPersonalExpenditure$Money_Spent[(USPersonalExpenditure$Category == "Personal Care") & (USPersonalExpenditure$Year == paste0("X", year))]), 
  Food_Tobacco = sum(USPersonalExpenditure$Money_Spent[(USPersonalExpenditure$Category == "Food and Tobacco") & (USPersonalExpenditure$Year == paste0("X", year))]), 
  Household_op = sum(USPersonalExpenditure$Money_Spent[(USPersonalExpenditure$Category == "Household Operation") & (USPersonalExpenditure$Year == paste0("X", year))]), 
  Priv_Ed = sum(USPersonalExpenditure$Money_Spent[(USPersonalExpenditure$Category == "Private Education") & (USPersonalExpenditure$Year == paste0("X", year))])
  )
  lowest_cat <- which.min(expenditure_sums1)
  return(c(lowest_cat, expenditure_sums1))
}

lowest_category("1960")

lowest_category("1945")

```

```{r}

```


Using your function, determine the lowest spending category of each year
Hint: use the `sapply()` function to apply your function to a vector of years

```{r}

```

``` {r personal-practice}
name <- c("John", "Adam", "Lili", "Billy", "Oumayma", "Phung", "Manuel")

age <- c(19, 20, 18, 26, 26, 21, 21)

in_MATH3302 <- c(TRUE, FALSE, FALSE, TRUE, TRUE, TRUE, TRUE)

my_df <- data.frame(name, age, in_MATH3302)

# as.factor can be used to store information when there are multiple rows with the same piece of information in a column

nrow(my_df)

colnames(my_df) <- c("Name", "Age", "in_MATH3302")

# rownames_to_columns(df, var) converts the row names to column names

glimpse(penguins)

# View(penguins)

my_df$Age

my_df[["Age"]]

my_df[4, "in_MATH3302"]

my_df[my_df["in_MATH3302"] == FALSE, ]
# above and below are same thing
my_df[my_df$in_MATH3302 == FALSE, ]

# exercise
pts <- c(1, 0, 3, 1)
pts_allowed <- c(5, 0, 2, 3)

games <- data.frame(pts, pts_allowed)

games$diff <- pts - pts_allowed

games$won <- games$diff > 0

games$opponents <- c("North Caroline Courage", "Racing Louisville", "Bay FC", "Washington Spirit")

games
```